Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-17 doc/lean.tex
General updates

We have implemented the rightmost two components of Figure~\ref{fig:chain}, namely the
proof checker and both model and weighted model counters, in the Lean 4 programming language
and proof assistant [cite].
In this section, we briefly describe the functions we have implemented in Lean and
the specifications we have proved.
More information is provided in the Appendix~\ref{appendix:lean}.

@@

Our specifications are built on a generic library for the syntax and semantics of
propositional logic. We define the data type \lstinline{PropForm Var} of propositional
formulas with variables \lstinline{Var}, which are positive natural numbers,
corresponding to the indexing of variables in the DIMACS format for SAT.
It is convenient to take a truth assignment \lstinline{PropAssignment Var} to a function that
assigns a boolean value to every assignment, and later restrict attention to subsets of the
variables.

@@

The goal of the proof checker is to construct a POG that is equivalent to the input CNF.
Each element of a POG  is either a variable, a binary disjunction,
or an arbitrary conjunction:
\begin{lstlisting}
inductive PogElt where
  | var  : Var → PogElt
  | disj : Var → ILit → ILit → PogElt
  | conj : Var → Array ILit → PogElt
\end{lstlisting}
In the first case, the argument \lstinline{x} in the expression
\lstinline{var x} is the index
of the variable; in \lstinline{disj x left right} and \lstinline{conj x args}
it is the definition number in the CRAT file. Note that representing edges as literals allows us to negate the arguments to \lstinline{disj} and \lstinline{conj}.
A \lstinline{Pog} is then simply an array of \lstinline{PogElt}.
For each POG \lstinline{P} and literal \lstinline{l}, we define
\lstinline{P.toPropForm l} to be the
propositional formula that arises from interpreting \lstinline{l} as a propositional
formula, unfolding all the defined conjunctions and disjunctions.
Our formulation of what it means for such a formula to be partitioned, corresponding
to the definition at the end of \cref{section:logical:foundations}, is as follows:
\begin{lstlisting}
def partitioned : PropForm Var → Prop
  | tr | fls | var _ => True
  | neg φ    => φ.partitioned
  | disj φ ψ => φ.partitioned ∧ ψ.partitioned ∧
                    ∀ τ, ¬ (φ.eval τ ∧ ψ.eval τ)
  | conj φ ψ => φ.partitioned ∧ ψ.partitioned ∧ (φ.vars ∩ ψ.vars = ∅)
\end{lstlisting}

@@

\subsection{The Model Counters}

We have formalized the central quantity (\ref{eqn:rep}) in the ring evaluation problem,
\cref{def:ring_evaluation}, as follows:
\begin{lstlisting}
def weightCount {R : Type} [CommRing R]
    (weight : Var → R) (φ : PropForm Var) (s : Finset Var) : R :=
  ∑ τ in models φ s, ∏ x in s, if τ x then weight x else 1 - weight x
\end{lstlisting}
Here \lstinline{R} is assumed to be a commutative ring.
The counting scheme of Proposition~\ref{prop:ring:eval} for partitioned formulas is expressed as follows:
\begin{lstlisting}
 def ringEval (weight : Var → R) : PropForm Var → R
  | tr       => 1
  | fls      => 0
  | var x    => weight x
  | neg φ    => 1 - ringEval weight φ
  | disj φ ψ => ringEval weight φ + ringEval weight ψ
  | conj φ ψ => ringEval weight φ * ringEval weight ψ
\end{lstlisting}
We arbitrarily assign 0 to implications and bi-implications;
recall that, by definition, these do not appear in partitioned formulas.
Proposition~\ref{prop:ring:eval} is then formalized as follows:
\begin{lstlisting}
theorem ringEval_eq_weightCount (weight : Var → R) {φ : PropForm Var} :
    partitioned φ → ringEval weight φ = weightCount weight φ (vars φ)
\end{lstlisting}

>>>>

We have implemented the rightmost two components of Figure~\ref{fig:chain}, namely the
proof checker and both model and weighted model counters, in the \lean{} programming language
and proof assistant~\cite{demoura:cade:2021}.
In this section, we briefly describe the functions we have implemented in Lean and
the specifications we have proved.
More information is provided in the Appendix~\ref{appendix:lean}.

@@

Our specifications are built on a generic library for the syntax and semantics of
propositional logic. We define the data type \lstinline{PropForm Var} of propositional
formulas with variables \lstinline{Var}, which are positive natural numbers,
corresponding to the indexing of variables in the CNF file.
It is convenient to take a truth assignment \lstinline{PropAssignment Var} to a function that
assigns a Boolean value to every assignment, and later restrict attention to subsets of the
variables.

@@

The goal of the proof checker is to construct a POG that is equivalent to the input CNF.
Each element of a POG  is either a variable, a binary disjunction,
or an arbitrary conjunction:
\begin{lstlisting}
inductive PogElt where
  | var  : Var → PogElt
  | disj : Var → ILit → ILit → PogElt
  | conj : Var → Array ILit → PogElt
\end{lstlisting}
In the first case, the argument \lstinline{x} in the expression
\lstinline{var x} is the index
of the variable; in \lstinline{disj x left right} and \lstinline{conj x args}
it is the definition number in the CRAT file. Note that representing edges as literals allows us to negate the arguments to \lstinline{disj} and \lstinline{conj}.
A \lstinline{Pog} is then simply an array of \lstinline{PogElt}.
For each POG \lstinline{P} and literal \lstinline{l}, we define
\lstinline{P.toPropForm l} to be the
propositional formula that arises from interpreting \lstinline{l} as a propositional
formula, unfolding all the defined conjunctions and disjunctions.
A partitioned-operation formula, as defined in \cref{section:logical:foundations}, is expressed as:
\begin{lstlisting}
def partitioned : PropForm Var → Prop
  | tr | fls | var _ => True
  | neg φ    => φ.partitioned
  | disj φ ψ => φ.partitioned ∧ ψ.partitioned ∧
                    ∀ τ, ¬ (φ.eval τ ∧ ψ.eval τ)
  | conj φ ψ => φ.partitioned ∧ ψ.partitioned ∧ (φ.vars ∩ ψ.vars = ∅)
\end{lstlisting}

@@

\subsection{The Model Counters}
\label{subsection:counting}

We have formalized the central quantity (\ref{eqn:rep}) in the ring evaluation problem,
\cref{def:ring_evaluation}, as follows:
\begin{lstlisting}
def weightCount {R : Type} [CommRing R]
    (weight : Var → R) (φ : PropForm Var) (s : Finset Var) : R :=
  ∑ τ in models φ s, ∏ x in s, if τ x then weight x else 1 - weight x
\end{lstlisting}
Here \lstinline{R} is assumed to be a commutative ring.
The counting scheme of Proposition~\ref{prop:ring:eval} for partitioned formulas is expressed as follows:
\begin{lstlisting}
 def ringEval (weight : Var → R) : PropForm Var → R
  | tr       => 1
  | fls      => 0
  | var x    => weight x
  | neg φ    => 1 - ringEval weight φ
  | disj φ ψ => ringEval weight φ + ringEval weight ψ
  | conj φ ψ => ringEval weight φ * ringEval weight ψ
\end{lstlisting}
%%We arbitrarily assign 0 to implications and bi-implications;
%%recall that, by definition, these do not appear in partitioned formulas.
Proposition~\ref{prop:ring:eval} is then formalized as follows:
\begin{lstlisting}
theorem ringEval_eq_weightCount (weight : Var → R) {φ : PropForm Var} :
    partitioned φ → ringEval weight φ = weightCount weight φ (vars φ)
\end{lstlisting}

### 2023-03-17 doc/leanappendix.tex
General updates

We have implemented the rightmost two components of Figure~\ref{fig:chain}, namely the
proof checker and both model and weighted model counters, in the Lean 4 programming language
and proof assistant [cite]. Lean 4 implements a logical foundation in which expressions
have a computational interpretation, and just as with other proof assistants like Isabelle and Coq,
a function defined within the formal system can be compiled to efficient code.
At the same time, we can state and prove claims about the function within the system, thereby
verifying that the functions compute the intended results.

@@

In this section, we describe the implementation in Lean and the specifications we have proved.
Lean 4's logical foundation describes a functional programming language with
inductive data types,
but our procedures also make use of calculations with natural numbers, integers, and rationals,
and data structures for arrays and hashmaps.
Lean's code extraction uses efficient but unverified versions of these,
and so in Subsection~\ref{subsection:verification:specifics} we clarify
what exactly we have verified and what we currently trust.

@@

\subsection{The Proof Checker}
\label{subsection:proof:checker}

@@

Lean 4 does not yet have an efficient implementation of rational arithmetic, and
we have not yet implemented a verified version of $\drational$,
defined in Section~\ref{sect:experimental}.
We have therefore implemented a separate procedure that carries out an integer
calculation of the number of models.
The number of models that a propositional formula has depends on what we take the
variables to be. For example, there is only one assignment to the variables
$\{ x, y \}$ that makes $x \wedge y$ true, but there are two assignments to the
variables $\{ x, y, z \}$ that make $x \wedge y$ true.
It is not hard to show that for partitioned formulas \lstinline{PropForm ν} whose
variables are among a finite set \lstinline{s} of variables of cardinality \lstinline{numVars},
we can count the number of models of the formula on the variables in \lstinline{s}
recursively as follows:
\begin{lstlisting}
def countModels (nVars : Nat) : PropForm ν → Nat
  | tr       => 2^nVars
  | fls      => 0
  | var _    => 2^(nVars - 1)
  | neg φ    => 2^nVars - countModels nVars φ
  | disj φ ψ => countModels nVars φ + countModels nVars ψ
  | conj φ ψ => countModels nVars φ * countModels nVars ψ / 2^nVars
\end{lstlisting}
We have also implemented an efficient counting function for POGs and proved that
it computes that same quantity on the corresponding formulas.
Finally,
We have proved that the calculation above really does
return the number of models:
\begin{lstlisting}
theorem countModels_eq_card_models {φ : PropForm ν} {s : Finset ν}
    (hvars : vars φ ⊆ s) (hpar :  partitioned φ) :
  countModels (card s) φ = card (models φ s)
\end{lstlisting}
In particular, taking \lstinline{s} to be exactly the variables of \lstinline{φ},
we have that the number of models on its variables is \lstinline{countModels φ (card (vars φ))}.

@@

We have specified the behavior of the proof checker in Section~\ref{subsection:proof:checker};
in this section we sketch the implementation.
Establishing its correctness was, by far, the most challenging part of the formalization
process.
For clarity, we describe the algorithm with ordinary mathematical notation rather than
Lean snippets.

@@

\subsection{Evaluation}

[TODO: Say something to establish that the procedure run in reasonable time and produce
useful results; or maybe this goes in the next section?]

>>>>

We have implemented the rightmost two components of Figure~\ref{fig:chain}, namely the
proof checker and both model and weighted model counters, in the \lean{} programming language
and proof assistant [cite]. \lean{} implements a logical foundation in which expressions
have a computational interpretation, and just as with other proof assistants like Isabelle and Coq,
a function defined within the formal system can be compiled to efficient code.
At the same time, we can state and prove claims about the function within the system, thereby
verifying that the functions compute the intended results.

@@

In this section, we describe the implementation in Lean and the specifications we have proved.
\lean{}'s logical foundation describes a functional programming language with
inductive data types,
but our procedures also make use of calculations with natural numbers, integers, and rationals,
and data structures for arrays and hashmaps.
Lean's code extraction uses efficient but unverified versions of these,
and so in Subsection~\ref{subsection:verification:specifics} we clarify
what exactly we have verified and what we currently trust.

@@

\subsection{The Proof Checker}
\label{appendix:proof:checker}

@@

\lean{} does not yet have an efficient implementation of rational arithmetic, and
we have not yet implemented a verified version of $\drational$,
defined in Section~\ref{sect:experimental}.
We have therefore implemented a separate procedure that carries out an integer
calculation of the number of models.
The number of models that a propositional formula has depends on what we take the
variables to be. For example, there is only one assignment to the variables
$\{ x, y \}$ that makes $x \wedge y$ true, but there are two assignments to the
variables $\{ x, y, z \}$ that make $x \wedge y$ true.
It is not hard to show that for partitioned formulas \lstinline{PropForm ν} whose
variables are among a finite set \lstinline{s} of variables of cardinality \lstinline{numVars},
we can count the number of models of the formula on the variables in \lstinline{s}
recursively as follows:
\begin{lstlisting}
def countModels (nVars : Nat) : PropForm ν → Nat
  | tr       => 2^nVars
  | fls      => 0
  | var _    => 2^(nVars - 1)
  | neg φ    => 2^nVars - countModels nVars φ
  | disj φ ψ => countModels nVars φ + countModels nVars ψ
  | conj φ ψ => countModels nVars φ * countModels nVars ψ / 2^nVars
\end{lstlisting}
We have also implemented an efficient counting function for POGs and proved that
it computes that same quantity on the corresponding formulas.
Finally,
We have proved that the calculation above really does
return the number of models:
\begin{lstlisting}
theorem countModels_eq_card_models {φ : PropForm ν} {s : Finset ν}
    (hvars : vars φ ⊆ s) (hpar :  partitioned φ) :
  countModels (card s) φ = card (models φ s)
\end{lstlisting}
In particular, taking \lstinline{s} to be exactly the variables of \lstinline{φ},
we have that the number of models on its variables is \lstinline{countModels φ (card (vars φ))}.

@@

We have specified the behavior of the proof checker in Section~\ref{appendix:proof:checker};
in this section we sketch the implementation.
Establishing its correctness was, by far, the most challenging part of the formalization
process.
For clarity, we describe the algorithm with ordinary mathematical notation rather than
Lean snippets.

@@

%% \subsection{Evaluation}

%% [TODO: Say something to establish that the procedure run in reasonable time and produce
%% useful results; or maybe this goes in the next section?]
